Component response rate variation underlies the stability of complex systems (revision notes)

### 2020-01-19 notebook/revision_notes.Rmd
Edit to fix compile error

In complex systems represented by large random matrices, correlation between matrix elements $A_{ij}$ and $A_{ji}$ affects the distribution of eigenvalues and therefore local stability. As the correlation between matrix elements ($\rho$) decreases, the eigenvalue spectra changes such that more variation falls along the imaginary axis. The figure panels below compare a random matrix in which $\rho = 0$ (left) to one in which $\rho$ = -0.5 (right). In both panels, complex systems include $S = 1000$ components, with diagonal elements of $-1$ and off-diagonal elements drawn from a normal distribution with a mean of $\mu = 0$ and standard deviation of $\sigma = 0.4$.

@@

When $\textbf{A}$ was stabilised by $Var(\gamma)$, the change in $\rho$ was normally distributed around `r mean(rho_change)` (95\% CIs: `r rho_change_CI[1]`, `r rho_change_CI[2]`). Hence, decreasing the correlation between $\textbf{A}_{ij}$ and $\textbf{A}_{ji}$ was not by itself the cause of stability. For a clearer picture of the effect of $\gamma$, it is useful to show the relationship between $\rho$ and $\max(\Re(\lambda))$ again as above, but this time also for how the relationship changes given $Var(\gamma)$. That is, given $\textbf{M} = \textbf{A}$ in addition to $\textbf{M} = \gamma\textbf{A}$.

@@

```{r, echo = FALSE}
mat_S16   <- build_rho_mat(S = 1600, sigma = (1/40), rho = 0);
mat_S16e  <- eigen(mat_S16)$values;
mat_S16r  <- Re(mat_S16e);
mat_S16i  <- Im(mat_S16e);
plot(x = mat_S16e, y = mat_S16i, pch = 4, cex = 0.7, col = "firebrick",
     xlab = "Real", ylab = "Imaginary", cex.lab = 1.25, cex.axis = 1.25, 
     asp = 1);
```

>>>>

In complex systems represented by large random matrices, correlation between matrix elements $A_{ij}$ and $A_{ji}$ affects the distribution of eigenvalues and therefore local stability. As the correlation between matrix elements ($\rho$) decreases, the eigenvalue spectra changes such that more variation falls along the imaginary axis. The figure panels below compare a random matrix in which $\rho = 0$ (left) to one in which $\rho = -0.5$ (right). In both panels, complex systems include $S = 1000$ components, with diagonal elements of $-1$ and off-diagonal elements drawn from a normal distribution with a mean of $\mu = 0$ and standard deviation of $\sigma = 0.4$.

@@

When $\textbf{A}$ was stabilised by $Var(\gamma)$, the change in $\rho$ was normally distributed around `r mean(rho_change) * 10^6`$\times 10^{-6}$ (95\% CIs: `r rho_change_CI[1] * 10^6`$\times 10^{-6}$, `r rho_change_CI[2] * 10^6`$\times 10^{-6}$). Hence, decreasing the correlation between $\textbf{A}_{ij}$ and $\textbf{A}_{ji}$ was not by itself the cause of stability. For a clearer picture of the effect of $\gamma$, it is useful to show the relationship between $\rho$ and $\max(\Re(\lambda))$ again as above, but this time also for how the relationship changes given $Var(\gamma)$. That is, given $\textbf{M} = \textbf{A}$ in addition to $\textbf{M} = \gamma\textbf{A}$.

@@

```{r, echo = FALSE}
mat_S16   <- build_rho_mat(S = 1600, sigma = (1/40), rho = 0);
mat_S16e  <- eigen(mat_S16)$values;
mat_S16r  <- Re(mat_S16e);
mat_S16i  <- Im(mat_S16e);
plot(x = mat_S16r, y = mat_S16i, pch = 4, cex = 0.7, col = "firebrick",
     xlab = "Real", ylab = "Imaginary", cex.lab = 1.25, cex.axis = 1.25, 
     asp = 1);
```

### 2020-01-19 notebook/revision_notes.tex
Edit to fix compile error

In complex systems represented by large random matrices, correlation
between matrix elements \(A_{ij}\) and \(A_{ji}\) affects the
distribution of eigenvalues and therefore local stability. As the
correlation between matrix elements (\(\rho\)) decreases, the eigenvalue
spectra changes such that more variation falls along the imaginary axis.
The figure panels below compare a random matrix in which \(\rho = 0\)
(left) to one in which \(\rho\) = -0.5 (right). In both panels, complex
systems include \(S = 1000\) components, with diagonal elements of
\(-1\) and off-diagonal elements drawn from a normal distribution with a
mean of \(\mu = 0\) and standard deviation of \(\sigma = 0.4\).

@@

When \(\textbf{A}\) was stabilised by \(Var(\gamma)\), the change in
\(\rho\) was normally distributed around 4.8394472\times 10\^{}\{-6\}
(95\% CIs: -7.0762643\times 10\^{}\{-5\}, 8.0991595\times 10\^{}\{-5\}).
Hence, decreasing the correlation between \(\textbf{A}_{ij}\) and
\(\textbf{A}_{ji}\) was not by itself the cause of stability. For a
clearer picture of the effect of \(\gamma\), it is useful to show the
relationship between \(\rho\) and \(\max(\Re(\lambda))\) again as above,
but this time also for how the relationship changes given
\(Var(\gamma)\). That is, given \(\textbf{M} = \textbf{A}\) in addition
to \(\textbf{M} = \gamma\textbf{A}\).

@@

\begin{verbatim}
## Warning in xy.coords(x, y, xlabel, ylabel, log): imaginary parts discarded
## in coercion
\end{verbatim}

@@

Hence, under the same set of parameters, we predict a linear effect of
\(\rho\) on \(\max(\Re(\lambda)) = 0\), such that
\(\max(\Re(\lambda)) = \rho\).

Allesina and Tang note that \(\rho\) is defined as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{1}},

@@

We can rewrite as the more typical way of writing covariance (top),

@@

>>>>

In complex systems represented by large random matrices, correlation
between matrix elements \(A_{ij}\) and \(A_{ji}\) affects the
distribution of eigenvalues and therefore local stability. As the
correlation between matrix elements (\(\rho\)) decreases, the eigenvalue
spectra changes such that more variation falls along the imaginary axis.
The figure panels below compare a random matrix in which \(\rho = 0\)
(left) to one in which \(\rho = -0.5\) (right). In both panels, complex
systems include \(S = 1000\) components, with diagonal elements of
\(-1\) and off-diagonal elements drawn from a normal distribution with a
mean of \(\mu = 0\) and standard deviation of \(\sigma = 0.4\).

@@

When \(\textbf{A}\) was stabilised by \(Var(\gamma)\), the change in
\(\rho\) was normally distributed around 4.8394472\(\times 10^{-6}\)
(95\% CIs: -73.6979659\(\times 10^{-6}\), 83.9237307\(\times 10^{-6}\)).
Hence, decreasing the correlation between \(\textbf{A}_{ij}\) and
\(\textbf{A}_{ji}\) was not by itself the cause of stability. For a
clearer picture of the effect of \(\gamma\), it is useful to show the
relationship between \(\rho\) and \(\max(\Re(\lambda))\) again as above,
but this time also for how the relationship changes given
\(Var(\gamma)\). That is, given \(\textbf{M} = \textbf{A}\) in addition
to \(\textbf{M} = \gamma\textbf{A}\).

@@

@@

Hence, under the same set of parameters, we predict a linear effect of
\(\rho\) on \(\max(\Re(\lambda)) = 0\), such that
\(\max(\Re(\lambda)) = \rho\). Allesina and Tang note that \(\rho\) is
defined as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{1}},

@@

Since we define \(\sigma = 1/\sqrt{S}\) and \(C = 1\),
\(\sigma\sqrt{SC} = 1\), so the distribution of eigenvalues is circular
with a radius of unity. Further, since we set \(diag(\textbf{M}) = -1\),
we know that \(E[\max(\Re(\lambda))] = 0\).

Gibbs et al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{2}}
showed analytically that as \(S \to \infty\), the effect of a diagonal
matrix \(\textbf{X}\) on the stability of \(\textbf{M} = \textbf{XA}\)
becomes negligible. They also note that local stability of a random
matrix can be predicted from simple moments affecting three parameters,

\[E_{1} = \frac{1}{S(1-S)}\sum_{i \neq j}A_{ij},\]

\[E_{2} = \frac{1}{S(1-S)}\sum_{i \neq j}A^{2}_{ij} - E^{2}_{1},\]

and,

\[E_{c} = \frac{1}{S(1-S)E^{2}_{2}}\sum_{i \neq j} A_{ij} A_{ji} - \frac{E^{2}_{1}}{E^{2}_{2}}.\]

Note that given \(\mu = 0\) and \(C = 1\), \(E_{1} = 0\), while
\(E^{2}_{2} = Var(A_{ij})\), and \(E_{c} = \rho\). This suggests that my
parameterisations should only be affected by \(Var(A_{ij})\) and
\(\rho\). This does indeed appear to be the case if we restrict
\(\sigma = 1/\sqrt{S}\), where if \(\rho = 0\), random matrices with
\(\gamma\) have similar \(E[\max(\Re(\lambda))]\) to those without. But
the relationship between \(\rho\) and \(E[\max(\Re(\lambda))]\) clearly
differs when \(\gamma\) is included, and as \(S\) increases such that
\(\sigma > 1/\sqrt{S}\).

Hence, assuming high \(S\) and \(\rho = 0\), we should not see an effect
of \(\gamma\) on \(E[\max(\Re(\lambda))]\). But there is clearly an
effect of \(\gamma\) for nonzero \(\rho\) such that
\(\partial E[\max(\Re(\lambda))] / \partial\rho > 0\) and
\(\partial^{2} E[\max(\Re(\lambda))] / \partial\rho > 0\).

Note that, by definition, we have not changed \(E[M_{ij}]\) with the
addition of \(\gamma\), since \(E[\gamma_{ii}] = 1\). Elements of
\(\gamma\) are pulled from a random uniform distribution between 0 and
2, meaning that no covariance should be induced betwen \(M_{ij}\) and
\(M_{ji}\).

@@

\hypertarget{ref-Gibbs2017}{}
2. Gibbs, T., Grilli, J., Rogers, T. \& Allesina, S. The effect of
population abundances on the stability of large random ecosystems.
\emph{Physical Review E - Statistical, Nonlinear, and Soft Matter
Physics} \textbf{98,} 022410 (2018).
